Variance reduction for MCMC methods via martingale representations

### 2020-03-04 aap/Revision/vr-mcmc-revision-new.tex
lemma 8 update

\subsection{Gaussian mixtures}
 We consider a sample generated by ULA with $\pi$  given by the mixture of two equally-weighted $d$-dimensional Gaussian distributions of the following form
 \[
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
 \]
where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{d \times d}$. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}, \quad
\]
\[
\nabla U(x) = \Sigma^{-1}(x-\mu) +2\mu^\top\Sigma^{-1}\left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1},
\]
respectively.
In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and $\Sigma = \Sigma_0$ with randomly generated positive-definite matrix $\Sigma_0$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent training trajectories, each one of size $n = 10^4$.  Then we solve the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 1$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and plot boxplots of the ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials, since they are used as regressors during computations of $\hat{a}_{p-l,\mathbf{k}}$.

@@

Now, from the H\"older inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
\begin{align*}
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
 & \leq\sum_{j=1}^{q} \left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
\end{align*}
Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma\mu'(X_{k-1})\right),$
we get
\[
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
\]
where we denote
\[
\zeta_{l,j}(u)=\mathsf{E}\left[e^{-u\sum_{k=l}^{j}\gamma\mu'(X_{k-1})}\right],\quad u>0.
\]
From Theorem~\ref{thm:ula_const_size} in Appendix~\ref{sec:appendix}, it follows that
\[
\mathsf{E}\left[e^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq e^{u^2\varkappa (j-l-1)\gamma^{2}}
\]
with constant $\varkappa = ...$, where we used the fact that $\mu^{\prime}(x)$ is bounded. Then we have
\[
\mathsf{E}\left[e^{-s\sum_{k=l}^{j}\gamma_{k}\mu'(X_{k-1})}\right]\leq e^{s^2\varkappa \gamma^2(j-l-1)}e^{-s\gamma (j-l-1)\pi_\gamma(\mu')}.
\]
\denis{Please finish}
Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$, we have
\[
\pi(\mu')=\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=2\int\mu^{2}(x)\pi(x)\,dx>0
\]
yielding the final bound
\[
\zeta_{l,j}(s) \leq e^{s^2\kappa (j-l-1) \gamma^2}e^{-2s\gamma(j-l-1)\alpha}
\]
where $ \alpha = \int \mu^2(x)\pi(x)\,dx$.
Finally
\begin{align*}
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]  \lesssim
\end{align*}
Plugging into bound for conditional variance,
\begin{align*}
\mathsf{Var}\left[\sum_{p=l}^{q}\gamma_{p+1}\partial_{y_{l}}^{K}f\left(X_{p}\right)\Big|X_{l-1}\right] \leq
\frac{A_{K}^{2}e^{2\kappa \Gamma^*}}{4\alpha^2(1-\gamma_1 B_\mu)^3}
\end{align*}
\par
The proof is completed.
\end{proof}

>>>>

\subsection{Gaussian mixtures}
 We consider a sample generated by ULA with $\pi$  given by the mixture of two equally-weighted $d$-dimensional Gaussian distributions of the following form
 \[
 \pi(x) = \frac{1}{2\sqrt{(2\pi)^{d}|\Sigma|}} \left( \rme^{-\frac{(x-\mu)^T\Sigma^{-1}(x-\mu)}{2}} + \rme^{-\frac{(x+\mu)^T\Sigma^{-1}(x+\mu)}{2}}\right)
 \]
where $\rho \in (0,1), \mu \in \rset^d, \Sigma \in \rset^{d \times d}$ and $|\Sigma|$ is its determinant. The function $U(x)$ and its gradient are given by
\[
U(x) = \frac{1}{2}(x-\mu)^T\Sigma^{-1}(x-\mu) - \ln{\left(1 + \rme^{-2\mu^\top \Sigma^{-1}x}\right)}, \quad
\]
\[
\nabla U(x) = \Sigma^{-1}(x-\mu) +2\mu^\top\Sigma^{-1}\left(1 + \rme^{2 \mu^\top\Sigma^{-1} x}\right)^{-1},
\]
respectively.
In our experiments we considered $d = 2$, $\mu = \left(0.5,0.5\right)$ and $\Sigma = \Sigma_0$ with randomly generated positive-definite matrix $\Sigma_0$ (heterogeneous structure). In order to approximate the expectation \(\pi(f)\) with \(f(x)=\sum_{i=1}^d x_i\) or \(f(x) = \sum_{i=1}^d x^2_i\) we used constant step size $\gamma=0.1$ and sampled $\NtrainPath = 10$ independent training trajectories, each one of size $n = 10^4$.  Then we solve the least squares  problems \eqref{eq:Qcircr} using the first order polynomial approximations for the coefficients \(a_{p,\mathbf{k}}\) as described in the previous section, hence we set $K = 1$. We set the truncation level \(n_{\mathrm{trunc}} = 50\). To test our variance reduction algorithm, we generated 100 independent trajectories of length $n = 10^3$ and plot boxplots of the ergodic averages in Figure ~\ref{fig:GMM}. Our approach (MDCV, Martingale decomposition control variates) is compared to other variance reduction methods of \cite{mira2013zero} and \cite{belomestny2019esvm}. In the baselines we use first order polynomials, since they are used as regressors during computations of $\hat{a}_{p-l,\mathbf{k}}$.

@@

Now, from the H\"older inequality, we obtain (with $\|X\|_p = (\mathsf{E}X^p)^{\frac{1}{p}}$)
\begin{align*}
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=l}^{j}\alpha_{k}\left( \sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right] & \leq\sum_{j=1}^{q}\left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left\Vert \sum_{p=j+1}^{q+1} \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}^{2}\\
 & \leq\sum_{j=1}^{q} \left\Vert \prod_{k=1}^{j}\alpha_{k}\right\Vert _{2}\left(\sum_{p=j+1}^{q+1}\left\Vert \prod_{k=j+1}^{p}\alpha_{k}\right\Vert _{4}\right)^{2}.
\end{align*}
Now using the fact that $\prod\limits_{k=j+1}^{p}\alpha_{k} \leq \exp\left(-\sum\limits_{k=j+1}^{p}\gamma\mu'(X_{k-1})\right),$
we get
\[
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]\leq\sum_{j=1}^{q} \zeta^{1/2}_{1,j}(2) \left(\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4)\right)^{2},
\]
where we denote
\[
\zeta_{l,j}(u)=\mathsf{E}\left[e^{-u\gamma\sum_{k=l}^{j}\mu'(X_{k-1})}\right],\quad u>0.
\]
From Theorem~\ref{thm:ula_const_size} in Appendix~\ref{sec:appendix}, it follows that
\[
\mathsf{E}\left[e^{-u\gamma\left(\sum_{k=l}^{j}[\mu'(X_{k-1})-\pi_\gamma(\mu')]\right)}\right]\leq e^{u^2\varkappa (j-l+1)\gamma^{2}}
\]
with constant $\varkappa = ...$, where we used the fact that $\mu^{\prime}(x)$ is bounded. Then we have
\[
\mathsf{E}\left[e^{-s\gamma\sum_{k=l}^{j}\mu'(X_{k-1})}\right]\leq e^{s^2\varkappa \gamma^2(j-l+1)}e^{-s\gamma (j-l+1)\pi_\gamma(\mu')}.
\]
Furthermore, since $\mu\pi',\mu'\pi\in L^{1}(\mathbb{R})$ and $\pi^{\prime}(x) = -\frac{1}{2}\pi(x)\mu(x)$, we have
\[
\pi(\mu')=\int\mu'(x)\pi(x)\,dx=-\int\mu(x)\pi'(x)\,dx=\frac{1}{2}\int\mu^{2}(x)\pi(x)\,dx>0
\]
Note also that \(\pi_\gamma(\mu^{\prime}) \geq \pi_\gamma(\mu^{\prime}) - C_1\sqrt{\gamma}\)
yielding the bound
\[
\zeta_{l,j}(s) \leq e^{s^2\kappa (j-l+1) \gamma^2}e^{-s\gamma(j-l+1)\alpha}
\]
where $ \alpha = \frac{1}{2}\int \mu^2(x)\pi(x)\,dx - C_1\sqrt{\gamma} > 0$ for sufficiently small $\gamma$. Hence we obtain
\begin{align*}
\sum_{p=j+1}^{q+1} \zeta^{1/4}_{j+1,p}(4) \leq \sum_{p=j+1}^{q+1}e^{4\kappa(p-j)\gamma^2}e^{-2\gamma(p-j)\alpha} < \frac{e^{4\kappa\gamma^2 - 2\gamma\alpha}}{1-e^{4\kappa\gamma^2 - 2\gamma\alpha}}
\end{align*}
and thus the final bound follows:
\begin{align*}
\mathsf{E}\left[\sum_{j=1}^{q}\prod_{k=1}^{j}\alpha_{k}\left(\sum_{p=j+1}^{q+1}\prod_{k=j+1}^{p}\alpha_{k}\right)^{2}\right]  \leq \frac{e^{6\kappa\gamma^2 - 3\gamma\alpha}}{\left(1-e^{4\kappa\gamma^2 - 2\gamma\alpha}\right)\left(1-e^{2\kappa\gamma^2 - \gamma\alpha}\right)}
\end{align*}
Plugging into variance bound,
\begin{align*}
\mathsf{Var}_x\left[\sum_{p=1}^{q}\partial_{y_{1}}^{K}f\left(X_{p}\right)\right] \leq
\frac{A_{K}^{2}e^{2\kappa \Gamma^*}}{4\alpha^2(1-\gamma_1 B_\mu)^3}
\end{align*}
\par
The proof is completed.
\end{proof}
